---
title: "Creating Questionnaire ADaM: Control of Eating (ADCOEQ)"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Questionnaire ADaM: Control of Eating (ADCOEQ)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

Note that University of Leeds are the copyright holders of the CoEQ and the test 
data included within {admiralmetabolic} as well as the ADCOEQ code are for 
not-for-profit use only within {admiralmetabolic} and pharmaverse-related 
examples/documentation. Any persons or companies wanting to use the CoEQ should 
request a license to do so from the following [link](https://licensing.leeds.ac.uk/product/control-of-eating-questionnaire-coeq).

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(admiral)
library(admiralmetabolic)
```

## Example Data

In this vignette we use example data for the CoEQ. 
The example `QS` data (`qs_metabolic`) is included in the admiralmetabolic package.

```{r}
qs <- admiralmetabolic::qs_metabolic
```

## QS:
```{r echo=FALSE}
dataset_vignette(qs)
```

```{r echo=FALSE}
dm <- admiralmetabolic::dm_metabolic

adsl <- dm %>%
  select(-DOMAIN) %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM) %>%
  left_join(admiral_adsl %>% select(USUBJID, TRTSDT, TRTEDT), by = "USUBJID")
```

## ADSL:
```{r echo=FALSE}
dataset_vignette(adsl)
```

# Original Items

The original items, i.e. the answers to the questionnaire questions, can be
handled in the same way as in a [BDS finding ADaM](bds_finding.html). For
example:

```{r eval=TRUE}
adcoeq1 <- qs %>%
  # Add ADSL variables
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = exprs(TRTSDT, TRTEDT, TRT01P, TRT01A),
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  # Add analysis parameter variables
  mutate(
    PARAMCD = QSTESTCD,
    PARAM = QSTEST,
    PARCAT1 = QSCAT
  ) %>%
  # Add timing variables
  derive_vars_dt(new_vars_prefix = "A", dtc = QSDTC) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT)) %>%
  mutate(
    AVISIT = if_else(ADT <= TRTSDT, "BASELINE", VISIT),
    AVISITN = if_else(ADT <= TRTSDT, 0, VISITNUM)
  ) 
```

If the numeric values of the answers in `QSSTRESN` values are not ready to be used for deriving
scores and require transformation, it is recommended that `QSSTRESN` is kept in
the ADaM dataset for traceability, and the transformed value is stored in
`AVAL`, since that's what will be used for the score calculation.
For the CoEQ item 6 has to be handled in such way:

```{r eval=TRUE}
adcoeq2 <- adcoeq1 %>%
  # Add analysis parameter variables
  mutate(
    AVAL = ifelse(PARAMCD=='COEQ06', 100-QSSTRESN, QSSTRESN),
    AVALC = ifelse(PARAMCD=='COEQ20', QSORRES, NA)
  ) 
```

```{r echo=FALSE}
dataset_vignette(
  arrange(adcoeq2, USUBJID, PARCAT1, ADY, PARAMCD),
  display_vars = exprs(USUBJID, PARAMCD, PARAM, PARCAT1, QSSTRESN, AVALC, AVAL, ADY, AVISIT)
)
```

We handle unscheduled visits as normal visits. For deriving visits based on
time-windows, see [Visit and Period Variables](visits_periods.html#visits). And
for flagging values to be used for analysis, see `derive_var_extreme_flag()`.


# Scales and Scores

Scales and Scores are often derived as the sum or the average across a subset of
the items. 
For the Control of Eating questionnaire, four subscales are derived:

* Craving Control: Calculate average of items 9, 10, 11, 12 and 19.

* Craving for Sweet: Calculate average of items 3, 13, 14 and 15.

* Craving for Savoury: Calculate average of items 4, 16, 17 and 18.

* Positive Mood: Calculate average of items 5, 7, 8 and 6 (reversed).

These parameters can be derived by `derive_summary_records()`:
```{r eval=TRUE}
adcoeq3 <- adcoeq2 %>%
  derive_summary_records(
    dataset = .,
    dataset_add = .,
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN, ADT, ADY, PARCAT1, TRTSDT, TRTEDT, TRT01P, TRT01A),
    filter_add = (PARAMCD %in% c('COEQ09','COEQ10','COEQ11','COEQ12','COEQ19')),
    set_values_to = exprs(
      AVAL = mean(AVAL, na.rm = TRUE),
      PARAMCD = "COEQCRCO",
      PARAM = "COEQ - Craving Control"
    )
  ) %>%
  derive_summary_records(
    dataset = .,
    dataset_add = .,
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN, ADT, ADY, PARCAT1, TRTSDT, TRTEDT, TRT01P, TRT01A),
    filter_add = (PARAMCD %in% c('COEQ03','COEQ13','COEQ14','COEQ15')),
    set_values_to = exprs(
      AVAL = mean(AVAL, na.rm = TRUE),
      PARAMCD = "COEQCRSW",
      PARAM = "COEQ - Craving for Sweet"
    )
  ) %>%
  derive_summary_records(
    dataset = .,
    dataset_add = .,
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN, ADT, ADY, PARCAT1, TRTSDT, TRTEDT, TRT01P, TRT01A),
    filter_add = (PARAMCD %in% c('COEQ04','COEQ16','COEQ17','COEQ18')),
    set_values_to = exprs(
      AVAL = mean(AVAL, na.rm = TRUE),
      PARAMCD = "COEQCRSA",
      PARAM = "COEQ - Craving for Savoury"
    )
  ) %>%
  derive_summary_records(
    dataset = .,
    dataset_add = .,
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN, ADT, ADY, PARCAT1, TRTSDT, TRTEDT, TRT01P, TRT01A),
    filter_add = (PARAMCD %in% c('COEQ05','COEQ07','COEQ08','COEQ06')),
    set_values_to = exprs(
      AVAL = mean(AVAL, na.rm = TRUE),
      PARAMCD = "COEQPOMO",
      PARAM = "COEQ - Positive Mood"
    )
  )
```

```{r echo=FALSE}
dataset_vignette(
  arrange(adcoeq3, USUBJID, ADY, PARAMCD),
  display_vars = exprs(USUBJID, PARAMCD, PARAM, AVAL, ADY, AVISIT)
)
```

After deriving the scores by visit, the baseline and change from baseline
variables can be derived:
```{r eval=TRUE}
adcoeq4 <- adcoeq3 %>%
  # Flag baseline records (last before treatement start)
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order = exprs(ADT),
      new_var = ABLFL,
      mode = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) %>%
  # Derive baseline and change from baseline variables
  derive_var_base(
    by_vars = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var = BASE
  ) %>%
  # Calculate CHG for post-baseline records
  # The decision on how to populate pre-baseline and baseline values of CHG is left to producer choice
  restrict_derivation(
    derivation = derive_var_chg,
    filter = AVISITN > 0
  ) %>%
  # Calculate PCHG for post-baseline records
  # The decision on how to populate pre-baseline and baseline values of PCHG is left to producer choice
  restrict_derivation(
    derivation = derive_var_pchg,
    filter = AVISITN > 0
  ) %>%
  # Derive sequence number
  derive_var_obs_number(
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(PARAMCD, ADT),
    check_type = "error"
  )
```

```{r echo=FALSE}
dataset_vignette(
  adcoeq4,
  display_vars = exprs(USUBJID, PARAMCD, PARAM, AVISIT, ADY, AVAL, BASE, CHG, PCHG)
)
```

