---
title: "Creating a Metabolic ADVS ADaM"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Metabolic ADVS ADaM}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating a vital signs ADaM for metabolic
clinical trials.

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around creating the general set-up of an `ADVS`
using `{admiral}` functions is the same. In this vignette, we focus on the most common
endpoints and their derivations mainly found in metabolic trials to
avoid repeating information and maintaining the same content in two
places. As such, the code in this vignette is not completely executable; we recommend consulting the ADVS template script to view the full workflow.

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiralmetabolic)
library(pharmaversesdtm)
library(dplyr)
```

# Programming Workflow

-   [Read in Data](#readdata)
-   [Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`](#paramcd)
-   [Derive Additional Parameters (e.g. `BMI` for `ADVS`)](#derive_param)
-   [Common Metabolic Endpoints](#common_endpoints)
    -   [Derive Categorization Variables (`AVALCATx`, `BASECATx`)](#cat)
    -   [Derive Criterion Variables (`CRITy`, `CRITyFL`, `CRITyFLN`)](#crit_vars)
-   [Remaining ADVS Set-up](#advs_end)

## Read in Data {#readdata}

To start, all data frames needed for the creation of `ADVS` should be
loaded into the global environment. Reading data will usually be a company specific process, however, for the purpose of this vignette, we will use example data from `{pharmaversesdtm}` and `{admiral}`. We will utilize `DM`, `VS` and `ADSL` for the basis of `ADVS`.

```{r, message=FALSE, warning=FALSE}
dm_metabolic <- admiralmetabolic::dm_metabolic
vs_metabolic <- admiralmetabolic::vs_metabolic
admiral_adsl <- admiral::admiral_adsl

dm <- convert_blanks_to_na(dm_metabolic)
vs <- convert_blanks_to_na(vs_metabolic)
admiral_adsl <- convert_blanks_to_na(admiral_adsl)
```

Within this vignette, `DM` is used as the basis for `ADSL`:

```{r eval=TRUE}
# Retrieve required variables from admiral ADSL for this vignette that are not present in DM dataset
adsl <- dm %>%
  select(-DOMAIN) %>%
  mutate(TRT01P = ARM, TRT01A = ACTARM) %>%
  left_join(admiral_adsl %>% select(USUBJID, TRTSDT, TRTEDT), by = "USUBJID")
```

The following steps are to merge `ADSL` variables with the source data and
derive the usual `ADVS` analysis variables. Note that only the sections required for this vignette are covered in the following steps. To get a detailed guidance on all the steps, refer the `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).

```{r, eval=TRUE, include=FALSE}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01P, TRT01A)

advs <- derive_vars_merged(
  vs,
  dataset_add = adsl,
  new_vars = adsl_vars,
  by_vars = exprs(STUDYID, USUBJID)
)

advs <- derive_vars_dt(advs, new_vars_prefix = "A", dtc = VSDTC)

advs <-
  derive_vars_dy(advs, reference_date = TRTSDT, source_vars = exprs(ADT))
```

## Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1` {#paramcd}

The next step is to assign parameter level values such as `PARAMCD`, `PARAM`,
`PARAMN`, `PARCAT1`, etc. For this, a lookup can be created based on
the SDTM `--TESTCD` value to join to the source data.
One key addition in metabolic trials are vital sign parameters associated to body measurements, such as `BMI`, `HIPCIR`, and `WSTCIR`.

```{r, echo=TRUE, message=FALSE}
param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM, ~PARAMN, ~PARCAT1, ~PARCAT1N,
  "HEIGHT", "HEIGHT", "Height (cm)", 1, "Anthropometric measurement", 1,
  "WEIGHT", "WEIGHT", "Weight (kg)", 2, "Anthropometric measurement", 1,
  "BMI", "BMI", "Body Mass Index(kg/m^2)", 3, "Anthropometric measurement", 1,
  "HIPCIR", "HIPCIR", "Hip Circumference (cm)", 4, "Anthropometric measurement", 1,
  "WSTCIR", "WSTCIR", "Waist Circumference (cm)", 5, "Anthropometric measurement", 1,
  "DIABP", "DIABP", "Diastolic Blood Pressure (mmHg)", 6, "Vital Sign", 2,
  "PULSE", "PULSE", "Pulse Rate (beats/min)", 7, "Vital Sign", 2,
  "SYSBP", "SYSBP", "Systolic Blood Pressure (mmHg)", 8, "Vital Sign", 2,
  "TEMP", "TEMP", "Temperature (C)", 9, "Vital Sign", 2
)
```

This lookup may now be joined to the source data and this is how the
parameters will look like:

```{r, eval=TRUE, include=TRUE, message=FALSE}
advs <- derive_vars_merged_lookup(
  advs,
  dataset_add = param_lookup,
  new_vars = exprs(PARAMCD, PARAM, PARAMN, PARCAT1, PARCAT1N),
  by_vars = exprs(VSTESTCD)
)
```

```{r, eval=TRUE, echo=FALSE}
advs_param <- distinct(advs, USUBJID, PARAMCD, VSTESTCD, PARAM, PARCAT1, PARCAT1N)

dataset_vignette(advs_param, display_vars = exprs(USUBJID, VSTESTCD, PARAMCD, PARAM, PARCAT1, PARCAT1N))
```


```{r, eval=TRUE, include=FALSE}
advs <- mutate(
  advs,
  AVAL = VSSTRESN
)
```

## Derive Additional Parameters (e.g. `BMI` for `ADVS`) {#derive_param}

In metabolic trials, `BMI` is often calculated at source. But while creating the `ADVS`
dataset, we re-derive `BMI` from the collected height and weight values.
This is done to ensure that the `BMI` is calculated consistently across
all subjects and visits.

In this step, we create parameter Body Mass Index (`BMI`) for the `ADVS`
domain using the `derive_param_bmi()` function. Note that only variables
specified in the `by_vars` argument will be populated in the newly
created records. Also note that if height is collected only once for a
subject use `constant_by_vars` to specify the function to merge by the
subject-level variable - otherwise BMI is only calculated for visits
where both are collected.

```{r eval=TRUE}
# Removing BMI collected at source from the dataset
advs <- advs %>% filter(!VSTESTCD == "BMI")

advs <- derive_param_bmi(
  advs,
  by_vars = c(
    get_admiral_option("subject_keys"),
    exprs(!!!adsl_vars, VISIT, VISITNUM, ADT, ADY, VSTPT, VSTPTNUM)
  ),
  set_values_to = exprs(
    PARAMCD = "BMI",
    PARAM = "Body Mass Index (kg/m^2)",
    PARAMN = 3,
    PARCAT1 = "Anthropometric measurement",
    PARCAT1N = 1
  ),
  get_unit_expr = VSSTRESU,
  filter = VSSTAT != "NOT DONE" | is.na(VSSTAT),
  constant_by_vars = exprs(USUBJID)
)
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  arrange(advs, USUBJID, VISITNUM, VSTPTNUM, ADT, PARAMCD),
  display_vars = exprs(USUBJID, VSTESTCD, PARAMCD, PARAM, VISIT, AVAL),
  filter = PARAMCD %in% c("BMI")
)
```

```{r eval=TRUE, include=FALSE}
advs <- restrict_derivation(
  advs,
  derivation = derive_var_extreme_flag,
  args = params(
    by_vars = c(get_admiral_option("subject_keys"), exprs(PARAMCD)),
    order = exprs(ADT, VSTPTNUM, VISITNUM),
    new_var = ABLFL,
    mode = "last"
  ),
  filter = (!is.na(AVAL) & ADT <= TRTSDT)
)

advs <- derive_var_base(
  advs,
  by_vars = c(get_admiral_option("subject_keys"), exprs(PARAMCD)),
  source_var = AVAL,
  new_var = BASE
)

advs <- derive_var_chg(advs)

advs <- derive_var_pchg(advs)
```

## Common Metabolic Endpoints {#common_endpoints}

In the following sections, we will explore some of the most common
endpoints typically observed in metabolic trials.

One such endpoint is the improvement in weight category from baseline to
the end of treatment, which is often assessed using Body Mass Index
(`BMI`). To capture this, we will derive variables such as `AVALCATy`
and `BASECATy`, as detailed in the subsequent section.

Additionally, the achievement of weight reduction thresholds, such as ≥
5%, ≥ 10%, or ≥ 15% from baseline to end of treatment or at a certain visit,
is a common endpoint in metabolic trials. To accommodate these criteria,
we will derive relevant criterion variables such as `CRITy`, `CRITyFL`,
and `CRITyFLN`, with the necessary functions provided by `{admiral}`
outlined below.

### Derive Categorization Variables (`AVALCATx`, `BASECATx`) {#cat}

`{admiral}` does not currently have a generic function to aid in
assigning `AVALCATy`/ `AVALCAvN` and `BASECATy`/ `BASECAvN` values.
Below is a simple example of how these values may be assigned:

For deriving categorization variables (`AVALCATx`, `BASECATx`)
`{admiral}` provides
[`derive_vars_cat()`](https://pharmaverse.github.io/admiral/dev/reference/derive_vars_cat.html) (see
documentation of the function for details).

```{r eval=TRUE}
avalcat_lookup <- exprs(
  ~PARAMCD, ~condition,                ~AVALCAT1,           ~AVALCA1N,
  "BMI",    AVAL < 18.5,               "Underweight",       1,
  "BMI",    AVAL >= 18.5 & AVAL < 25,  "Normal weight",     2,
  "BMI",    AVAL >= 25 & AVAL < 30,    "Overweight",        3,
  "BMI",    AVAL >= 30 & AVAL < 35,    "Obesity class I",   4,
  "BMI",    AVAL >= 35 & AVAL < 40,    "Obesity class II",  5,
  "BMI",    AVAL >= 40,                "Obesity class III", 6,
  "BMI",    is.na(AVAL),               NA_character_,       NA_integer_
)

# Derive BMI class (AVALCAT1, AVALCA1N)
advs <- advs %>%
  derive_vars_cat(
    definition = avalcat_lookup,
    by_vars = exprs(PARAMCD)
  )
```

```{r eval=TRUE, echo=FALSE}
dataset_vignette(
  advs,
  display_vars = exprs(USUBJID, PARAMCD, VISIT, AVAL, AVALCA1N, AVALCAT1),
  filter = PARAMCD == "BMI"
)
```

In a similar way, we will create `BASECATy`/ `BASECAvN` variables.

```{r eval=TRUE}
basecat_lookup <- exprs(
  ~PARAMCD, ~condition,                ~BASECAT1,           ~BASECA1N,
  "BMI",    BASE < 18.5,               "Underweight",       1,
  "BMI",    BASE >= 18.5 & BASE < 25,  "Normal weight",     2,
  "BMI",    BASE >= 25 & BASE < 30,    "Overweight",        3,
  "BMI",    BASE >= 30 & BASE < 35,    "Obesity class I",   4,
  "BMI",    BASE >= 35 & BASE < 40,    "Obesity class II",  5,
  "BMI",    BASE >= 40,                "Obesity class III", 6,
  "BMI",    is.na(BASE),               NA_character_,       NA_integer_
)

# Derive baseline BMI class (BASECAT1, BASECA1N)
advs <- advs %>%
  derive_vars_cat(
    definition = basecat_lookup,
    by_vars = exprs(PARAMCD)
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  advs,
  display_vars = exprs(USUBJID, PARAMCD, VISIT, AVAL, BASE, ABLFL, BASECA1N, BASECAT1),
  filter = PARAMCD == "BMI"
)
```

### Derive Criterion Variables (`CRITy`, `CRITyFL`, `CRITyFLN`) {#crit_vars}

For deriving criterion variables (`CRITy`, `CRITyFL`, `CRITyFLN`)
`{admiral}` provides
[`derive_vars_crit_flag()`](https://pharmaverse.github.io/admiral/dev/reference/derive_vars_crit_flag.html).
It ensures that they are derived in an ADaM-compliant way (see
documentation of the function for details).

In most cases the criterion depends on the parameter and in this case the higher order function [`restrict_derivation()`](https://pharmaverse.github.io/admiral/dev/reference/restrict_derivation.html) can be useful.
In the following example, the criterion flags for weight based on percentage change in weight reduction from baseline is derived. Additional criterion flags can be added as needed.

```{r eval=TRUE}
advs <- advs %>%
  restrict_derivation(
    derivation = derive_vars_crit_flag,
    args = params(
      condition = PCHG <= -5 & PARAMCD == "WEIGHT",
      description = "Achievement of ≥ 5% weight reduction from baseline",
      crit_nr = 1,
      values_yn = TRUE,
      create_numeric_flag = FALSE
    ),
    filter = VISITNUM > 0 & PARAMCD == "WEIGHT"
  ) %>%
  restrict_derivation(
    derivation = derive_vars_crit_flag,
    args = params(
      condition = PCHG <= -10 & PARAMCD == "WEIGHT",
      description = "Achievement of ≥ 10% weight reduction from baseline",
      crit_nr = 2,
      values_yn = TRUE,
      create_numeric_flag = FALSE
    ),
    filter = VISITNUM > 0 & PARAMCD == "WEIGHT"
  )
```

```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  arrange(advs, USUBJID, VISITNUM, VSTPTNUM, PARAMCD),
  display_vars = exprs(USUBJID, PARAMCD, PCHG, CRIT1, CRIT1FL, CRIT2, CRIT2FL),
  filter = PARAMCD %in% c("WEIGHT")
)
```

## Remaining ADVS Set-up {#advs_end}

The `{admiral}` [Creating a BDS Finding ADaM
vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html)
covers all the steps that are not shown here, such as merging the
parameter-level values, timing variables, and analysis flags.

# Example Scripts {#example}

ADaM | Sample Code
---- | --------------
ADVS | [ad_advs.R](https://github.com/pharmaverse/admiralmetabolic/blob/main/inst/templates/ad_advs.R){target="_blank"} 
